\textsc{PAPERCLIP}: Associating Astronomical Observations and Natural Language with Multi-Modal Models

### 2024-03-07 paper/main.tex

\begin{abstract}
We present PAPERCLIP (Proposal Abstracts Provide an Effective Representation for Contrastive Language-Image Pre-training), a method which associates astronomical observations imaged by surveys and telescopes with natural language  using a neural network model. The model is fine-tuned from a pre-trained Contrastive Language–Image Pre-training (CLIP) model using successful observing proposal abstracts, optionally summarized via guided generation using large language models (LLMs), and corresponding downstream observations. Using observations from the \hubble Space Telescope (HST) as an example, we show that the fine-tuned model embodies a meaningful joint representation between observations and text through tests targeting image retrieval (i.e., querying the closest observations using natural language queries) and description retrieval (i.e., querying the astrophysical object classes and science use cases closest to a given observation). Our study demonstrates the potential for using generalist rather than task-specific models for finding patterns in astronomical data, in particular by leveraging text as an interface. \githubmaster
\end{abstract}

@@

Machine learning (ML) is starting to have a significant impact in the sciences, with astrophysics being no exception.
%
ML methods have demonstrated promise at every stage of the research pipeline, from instrument design, to data acquisition, to its analysis \citep{huertas2022dawes}.
%
Until recently, most applications of ML within astrophysics have focused on augmenting traditional techniques in order to improve performance on specific tasks.
%
The {Foundation Model} paradigm, in contrast, seeks to develop generalist models which can be deployed to simultaneously tackle a wide range of tasks \citep{bommasani2021opportunities}.
%
% The paradigm has been highly successful in domains like computer vision and natural language processing, as demonstrated by the widespread use and adoption of tools like ChatGPT, Dall-E, Stable Diffusion, and CLIP.
%
% At the heart of the paradigm lies the triumph of scale -- scaling up model size, dataset size, and compute.
% %
These models are typically pre-trained on massive amounts of unlabeled data using self-supervised or weakly-supervised learning techniques, enabling them to learn powerful representations which can be used downstream.
%
Foundation models can often benefit from additional training ({fine tuning}) using a relatively small amounts of domain-specific data in order to increase their usefulness when applied to specialized domains.

There is considerable interest in developing custom foundation models for the sciences \citep[e.g., ][]{batatia2023foundation,subramanian2023towards}, with astrophysics being ripe for such an effort given the availability of large amounts of publicly-available data and diverse modes of interacting with it.
%
The multi-modality inherent to astrophysical observations, with different types of data (e.g., images, spectra, light curves, textual descriptions) often available for a given target object, presents a unique opportunity.
%
This multi-modality was recently exploited in \textsc{AstroCLIP}~\citep{lanusse2023astroclip} to obtain a joint embedding space between multi-band images and optical spectra from the Dark Energy Spectroscopic Instrument (DESI).
%
\textsc{AstroLLaMA}~\citep{nguyen2023astrollama,perkowski2024astrollama} is another recent effort to fine-tune a publicly-available model (\textsc{Llama-2}) on astrophysics-specific textual data from the arXiv.

In this paper we describe \text{PAPERCLIP} (Proposal Abstracts Provide an Effective Representation for Contrastive Language-Image Pre-training\footnote{Technically we fine tune rather than pre train, but ``PAPERCLIFT'' was rejected by the senior author of this paper.}), a method that connects astronomical image observations with natural language by leveraging the association between abstracts of successful observing proposals and images corresponding to downstream observations. 
%
Concretely, here we showcase the method using observations imaged by the \hubble Space Telescope (HST).
%
We show that fine-tuning a pre-trained CLIP ~\citep[Contrastive Language-Image Pre-training; ][]{radford2021learning} image-text model on observation-abstract associations results in meaningful joint representations through quantitative and qualitative evaluation tests.
%
The method opens up the possibility of interacting with astronomical survey data using free-form natural language as an interface, which is a cornerstone of the success of the modern foundation model paradigm.
%

@@

We curate a dataset of images of \hubble Space Telescope (HST) observations and corresponding text descriptions from publicly-available sources.
%
We rely on proposal abstracts from the Proposal Abstracts Catalog\footnote{\url{https://archive.stsci.edu/hst/proposal_abstracts.html}} -- a catalog of successful HST proposals -- to generate captions for the observations, optionally summarizing the abstracts via guided generation using LLMs (described in Sec.~\ref{sec:summarization} below).
%
The HST has been operational since its launch on April 24, 1990, and we use available proposals and observations up to the Cycle 30 science program, which commenced data-taking in 2022.
%

@@

\begin{table}[h!]
  \centering
  \begin{tabular}{m{0.21\textwidth} p{1.9cm} p{1.9cm} m{8cm}}
      \toprule
      \centering \bfseries \hubble image & \centering \bfseries Obs. cycle \\ (Year) & \centering \bfseries Prop. ID & \centering \bfseries Proposal abstract (clipped) \tabularnewline
      \midrule
      \centering \includegraphics[width=0.18\textwidth]{\thedatafolder/img_2.pdf} & \centering \input{\thedatafolder/cycle_2.txt} \\ (1999) & \centering \input{\thedatafolder/id_2.txt} &  {\scriptsize \input{\thedatafolder/abs1_2.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.18\textwidth]{\thedatafolder/img_1.pdf} & \centering \input{\thedatafolder/cycle_1.txt} \\ (2013) & \centering \input{\thedatafolder/id_1.txt} &  {\scriptsize \input{\thedatafolder/abs1_1.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.18\textwidth]{\thedatafolder/img_3.pdf} & \centering \input{\thedatafolder/cycle_3.txt} \\ (2016) & \centering \input{\thedatafolder/id_3.txt} &  {\scriptsize \input{\thedatafolder/abs1_3.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.18\textwidth]{\thedatafolder/img_0.pdf} & \centering \input{\thedatafolder/cycle_0.txt} \\ (2019) & \centering \input{\thedatafolder/id_0.txt} &  {\scriptsize \input{\thedatafolder/abs1_0.txt}} \tabularnewline
      \bottomrule
  \end{tabular}
  \caption{Examples of \hubble images (left-most column) and corresponding clipped proposal abstracts (right-most column). The observation cycle and corresponding year, as well as proposal ID are shown in the second and third columns, respectively. The proposal ID links to the Mikulski Archive for Space Telescopes (MAST) page corresponding to the proposal.}
  \label{tab:dataset}
\end{table}

@@

The guided generation process ensures that, in this case, the output of the LLM strictly conforms to the format of the following example:
% \SM{Have it be centered}
% \begin{lstlisting}[language=Python]
% {
%   'objects_and_phenomena': ['star forming galaxy', 'lensed galaxy', ...], 
%   'science_use_cases': ['measure lensing magnification', 'probe spectral energy distributions', ...]
% }
% \end{lstlisting}
\begin{center}
  \begin{jsoncode}
    \centering
          \color{black}\{
            \color{codegreen}'objects_and_phenomena'\color{black}: [\color{codegreen}'star forming galaxy', 'lensed galaxy'\color{black}], 
            \color{codegreen}'science_use_cases'\color{black}: [\color{codegreen}'measure lensing magnification'\color{black}]
          \color{black}\}
  \end{jsoncode}
  \end{center}
which is then used to construct the summarized caption by combining the two key elements.
%
Examples of raw abstract snippets and corresponding LLM-generated summaries are shown in Tab.~\ref{tab:datasetsumm}.
%
% We emphasize that the intended goal of summarization-via-guided-generation is to increase the signal between text and images by standardizing the captions used for fine-tuning the CLIP model, and compare the quantitative performance of the model vs using the raw abstracts in Sec.~\ref{sec:results}.
%
We train separate models using the raw abstracts and the LLM-generated summaries, and compare their performance on downstream tasks in Sec.~\ref{sec:results}.
%
We note that, even after summarization, the association signal is expected to be noisy, since aspects of the summarized caption may not be directly descriptive of the observed images.

@@

CLIP \citep[Contrastive Language-Image Pre-training;][]{radford2021learning} is a multi-modal neural network model pre-trained on a large corpus of image-text pairs via weak supervision using a contrastive loss.
%
Given a minibatch $\mathcal{B}$ of $|\mathcal{B}|$ image-text pairs $\{(I_i, T_i)\}$, the goal is to align the learned representations of corresponding (positive) pairs $(I_i, T_i)$ while repelling the representations of unaligned (negative) pairs $(I_i, T_{j\neq i})$.
%
Image and text encoders $f: I \rightarrow \mathbb R^{n_\text{emb}}$ and $g: T \rightarrow \mathbb R^{n_\text{emb}}$ are used to map images and text to a common embedding space of dimension $n_\text{emb}$.
%
We use the standard softmax-based bidirectional variant of the InfoNCE~\citep{oord2018representation} contrastive loss function introduced for training CLIP-style architectures \citep{radford2021learning}
%
\begin{equation}
  \label{eq:softmax_loss}
  \mathcal{L}(\mathcal{B})=-\frac{1}{2|\mathcal{B}|} \sum_{i=1}^{|\mathcal{B}|}\left(\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_i \cdot y_j / \tau}}+\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_j \cdot y_i / \tau}}\right)
\end{equation}
%
where ${x}_i={f\left(I_i\right)}/{\left\|f\left(I_i\right)\right\|}$ and ${y}_i={g\left(T_i\right)}/{\left\|g\left(T_i\right)\right\|}$ are the normalized representations of the $i$-th image and text, respectively, and $\tau$ is a learnable temperature hyperparameter.
%
Note that this loss treats the image and text representations symmetrically, ensuring that the two modalities are on the same footing.

We use the CLIP-ViT-B/16\footnote{\url{https://huggingface.co/openai/clip-vit-base-patch16}} \citep{radford2021learning} variant as the base pre-trained CLIP model.
%
This model uses a 12-layer, 12-head, 768-embedding dimension vision transformer as the image encoder and a 12-layer, 8-head, 512-embedding dimension sequence transformer as the text backbone.
%
The text encoder has a maximum length of 77 tokens and the image encoder a native resolution of $224\times224$ pixels.
%
Linear projection layers map the outputs of the image and text encoders to a common embedding space of dimension $n_\text{emb}=512$.
%
In total, the model has 149,620,737 trainable parameters.
%
This model was originally pre-trained on 400 million image-text pairs from internet data \citep{radford2021learning}.
%

@@

The base CLIP model is fine-tuned using the dataset described in Sec.~\ref{sec:dataset}, using either the LLM-summarized abstracts or raw proposal abstracts.
%
When using raw proposal abstracts, random chunks of the text delimited by periods are selected on the fly to fit within the maximum token length of the text encoder.
%
Images are augmented via random four-fold rotations (increments of $90^\circ$) and randomly cropped to the native resolution of the image encoder, maintaining $\sim 20\%$ of the area of the original image, at each training step.
%
Given the relatively modest size of the fine-tuning dataset, a batch size $|\mathcal B| = 32$ is used throughout; larger batch sizes were observed to be susceptible to overfitting.
%
The temperature hyperparameter $\tau$ was initialized to its pre-trained value.
%
We emphasize that the positive and negative image-text association is noisy and imperfect, since multiple images can be associated with the same abstract, and the goal of the fine-tuning process is to leverage the signal contained in this noisy association.

@@

Figure~\ref{fig:sim_valtrain} (right) shows the retrieval accuracy, as defined in Eq.~\eqrefb{eq:retrieval_accuracy}, as a function of the retrieval fraction $k$.
%
In this case, we evaluate all four models (fine tuned on raw abstracts (blue), fine-tuned on LLM-summarized abstracts (red), trained on LLM-summarized abstracts from scratch (yellow), and the base model (purple)) on the same captions dataset, the summarized abstracts, for a direct comparison.
%
Remarkably, the model trained on raw abstracts shows very similar performance when evaluated on the summarized abstracts compared to that trained on the summarized abstracts themselves, indicating that \emph{(1)} the image-text association signal is preserved in the summarization process, and \emph{(2)} the model is able to effectively leverage meaningful concepts in the noisy raw abstracts through weak supervision.

@@

Having aligned the image and text representations, we can embed a natural language query using the model and show the closest images by embedding when ranked by cosine similarity.
%
We show these in Tabs.~\ref{tab:tti_base} and \ref{tab:tti} for the base and fine-tuned models respectively using four simple curated queries: \texttt{dwarf galaxy} (small galaxies that typically orbit larger galaxies like the Milky Way), \texttt{Jupiter},  \texttt{SN1987A} (a specific, prominent supernova), and \texttt{strong lensing} (the phenomenon of bending of light due to the gravitational influence of a foreground distribution of matter). The proposal ID corresponding to the retrieved images is shown below each image, which contains a hyperlink to the MAST page corresponding to the proposal.

@@

The model fine-tuned with summarized abstracts (Tab.~\ref{tab:tti}), meanwhile, shows strikingly different behavior.
%
% For example, it is able to return images with processing and assembly artifacts particular to HST (e.g., the lines through the middle in some images), which typically receive low similarity scores when using the base model.
%
Images of Jupiter are returned for the \texttt{Jupiter} query, with the planet clearly visible in all four images.
%
The \texttt{dwarf galaxy}-queried images correspond to proposals aiming to measure the kinematics of the stellar cores of dwarf galaxies.
%
Supernova SN1987 itself can be seen in the three closest images for the \texttt{SN1987A} query with the fourth image being a supernova remnant.
%
Cluster-scale as well as galaxy-scale gravitational lenses are returned by the \texttt{strong lensing} query, with lensing patterns visible in the images.

@@

The second row (an image of supernova 1987A) highlights an interesting pattern -- the base model erroneously attributes the object at the center of the image to a gravitational lens, while the fine-tuned model correctly identifies it as a supernova remnant.

Note that we chose to illustrate qualitative performance on text and image retrieval using the model fine-tuned on summarized abstracts, rather than raw abstracts. We show analogous results for the model fine-tuned on raw abstracts in App.~\ref{app:eval_raw}. Although the two models show very similar quantitative performance on retrieval metrics (as shown in Fig.~\ref{fig:sim_valtrain}), they exhibit characteristic behavior in terms of objects (images/text) retrieved. We emphasize that for scientific usefulness, the most useful goal is not necessarily to retrieve the objects with the highest similarity scores, but rather to identify a diverse set of interesting candidates for manual follow-up and further analysis; both models are seen to perform sensibly, even if differently, in this regard.

@@

\begin{table}[h!]
  \centering
  \renewcommand{\arraystretch}{0.1}
  \begin{tabular}{m{3cm} m{3.6cm} m{3.6cm} m{4cm}}
      \toprule
      \centering \bfseries \hubble image & \centering \textbf{Top-4 text} \\ \small\textbf{\textcolor{deeppurple}{(base)}} & \centering  \textbf{Top-4 text} \\ \small\textbf{\textcolor{deepred}{(summary fine-tuned)}} & \centering \textbf{Summarized abstract} \\ (objects; `ground truth') \tabularnewline
      \midrule
      \centering \includegraphics[width=0.15\textwidth]{\thedatafolder/img_itt_0.pdf} & \centering \scriptsize \verbatiminput{\thedatafolder/sci_itt_base_0.txt} & \centering  \scriptsize \verbatiminput{\thedatafolder/sci_itt_0.txt} &  {\scriptsize \input{\thedatafolder/abs_itt_0.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.15\textwidth]{\thedatafolder/img_itt_1.pdf} & \centering \scriptsize \verbatiminput{\thedatafolder/sci_itt_base_1.txt} & \centering  \scriptsize \verbatiminput{\thedatafolder/sci_itt_1.txt} &  {\scriptsize \input{\thedatafolder/abs_itt_1.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.15\textwidth]{\thedatafolder/img_itt_2.pdf} & \centering \scriptsize \verbatiminput{\thedatafolder/sci_itt_base_2.txt} & \centering  \scriptsize \verbatiminput{\thedatafolder/sci_itt_2.txt} &  {\scriptsize \input{\thedatafolder/abs_itt_2.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.15\textwidth]{\thedatafolder/img_itt_3.pdf} & \centering \scriptsize \verbatiminput{\thedatafolder/sci_itt_base_3.txt} & \centering  \scriptsize \verbatiminput{\thedatafolder/sci_itt_3.txt} &  {\scriptsize \input{\thedatafolder/abs_itt_3.txt}} \tabularnewline
      \bottomrule
  \end{tabular}
  \caption{Text snippets from a curated list most closely matching a given image query (left-most column) by cosine similarity, shown for the \textcolor{deeppurple}{base} (CLIP-ViT-B/16) and \textcolor{deepred}{summary fine-tuned} models. The `ground truth' LLM-summarized abstract is shown in the right-most column.}
  \label{tab:itt}
\end{table}

@@

In this paper we present \textsc{PAPERCLIP}, a method for training domain-specific multi-modal models for astrophysics that associates observations imaged by telescopes with natural language in a common, semantically-meaningful embedding space.
%
We showcase an application to \hubble Space Telescope (HST) observations, where the model is fine-tuned from a pre-trained CLIP model using (optionally summarized) abstracts of successful \hubble proposals, leveraging a noisy association signal between text and images.
%
We show that \textsc{PAPERCLIP} significantly outperforms the base CLIP model in quantitative metrics, such as retrieval accuracy, as well as quality of text-to-image and image-to-text retrieval.
%
We also introduce a novel LLM summarization process which leverages guided generation to distill the content of proposal abstracts while preserving salient information. 
%
Overall, the procedure demonstrates the efficacy on fine-tuning generalist pre-trained models on small amounts of domain-specific data, in particular astronomical datasets.

>>>>

\begin{abstract}
We present PAPERCLIP (Proposal Abstracts Provide an Effective Representation for Contrastive Language-Image Pre-training), a method which associates astronomical observations imaged by surveys and telescopes with natural language  using a neural network model. The model is fine-tuned from a pre-trained Contrastive Language–Image Pre-training (CLIP) model using successful observing proposal abstracts, optionally summarized via guided generation using large language models (LLMs), and corresponding downstream observations. Using observations from the \hubble Space Telescope (HST) as an example, we show that the fine-tuned model embodies a meaningful joint representation between observations and text through tests targeting image retrieval (i.e., finding the most relevant observations using natural language queries) and description retrieval (i.e., querying for astrophysical object classes and science use cases most relevant to a given observation). Our study demonstrates the potential for using generalist rather than task-specific models for finding patterns in astronomical data, in particular by leveraging text as an interface. \githubmaster
\end{abstract}

@@

Machine learning (ML) is starting to have a significant impact in the sciences, with astrophysics being no exception.
%
ML methods have demonstrated promise at every stage of the research pipeline, from instrument design, to data acquisition, to its analysis \citep{huertas2022dawes}.
%
Until recently, most applications of ML within astrophysics have focused on augmenting traditional techniques in order to improve performance on specific tasks.
%
The {Foundation Model} paradigm, in contrast, seeks to develop generalist models which can be deployed to simultaneously tackle a wide range of tasks \citep{bommasani2021opportunities}.
%
% The paradigm has been highly successful in domains like computer vision and natural language processing, as demonstrated by the widespread use and adoption of tools like ChatGPT, Dall-E, Stable Diffusion, and CLIP.
%
% At the heart of the paradigm lies the triumph of scale -- scaling up model size, dataset size, and compute.
% %
These models are typically pre-trained on massive amounts of unlabeled data using self-supervised or weakly-supervised learning techniques, enabling them to learn powerful representations which can then be used downstream in different ways.
%
Foundation models can often benefit from additional training ({fine tuning}) using a relatively small amounts of domain-specific data in order to increase their usefulness when applied to specialized domains.

There is considerable interest in developing custom foundation models for the sciences \citep[e.g., ][]{batatia2023foundation,subramanian2023towards}, with astrophysics being ripe for such an effort given the availability of large amounts of publicly-available data and diverse modes of interacting with it.
%
The multi-modality inherent to astrophysical observations, with different types of data (e.g., images, spectra, light curves, textual descriptions) often available for a given target object, presents a unique opportunity.
%
This multi-modality was recently exploited in \textsc{AstroCLIP}~\citep{lanusse2023astroclip} to construct a joint embedding space between multi-band images and optical spectra from the Dark Energy Spectroscopic Instrument (DESI).
%
\textsc{AstroLLaMA}~\citep{nguyen2023astrollama,perkowski2024astrollama} is another recent effort to fine-tune a publicly-available model (\textsc{Llama-2}) on astrophysics-specific textual data from the arXiv.

In this paper we describe \text{PAPERCLIP} (Proposal Abstracts Provide an Effective Representation for Contrastive Language-Image Pre-training\footnote{Technically we fine tune rather than pre train, but ``PAPERCLIFT'' was rejected by the senior author of this paper.}), a method that connects astronomical image observations with natural language by leveraging the association between abstracts of successful observing proposals and images corresponding to downstream observations. 
%
Concretely, here we showcase the method using observations imaged by the \hubble Space Telescope (HST).
%
We show that fine-tuning a pre-trained CLIP ~\citep[Contrastive Language-Image Pre-training; ][]{radford2021learning} image-text model on observation-abstract pairs results in meaningful joint representations through quantitative and qualitative evaluation tests.
%
The method opens up the possibility of interacting with astronomical survey data using free-form natural language as an interface, which is a cornerstone of the success of the modern foundation model paradigm.
%

@@

We curate a dataset of \hubble Space Telescope (HST) image observations and corresponding text descriptions from publicly available sources.
%
We rely on proposal abstracts from the Proposal Abstracts Catalog\footnote{\url{https://archive.stsci.edu/hst/proposal_abstracts.html}} -- a catalog of successful HST proposals -- to generate captions for the observations, optionally summarizing them via guided generation using LLMs (described in Sec.~\ref{sec:summarization} below).
%
The HST has been operational since its launch on April 24, 1990, and we use available proposals and observations up to the Cycle 30 science program, which commenced data-taking in 2022.
%

@@

\begin{table}[h!]
  \centering
  \begin{tabular}{m{0.21\textwidth} p{1.9cm} p{1.9cm} m{8cm}}
      \toprule
      \centering \bfseries \hubble image & \centering \bfseries Obs. cycle \\ (Year) & \centering \bfseries Prop. ID & \centering \bfseries Proposal abstract (clipped) \tabularnewline
      \midrule
      \centering \includegraphics[width=0.18\textwidth]{\thedatafolder/img_2.pdf} & \centering \input{\thedatafolder/cycle_2.txt} \\ (1999) & \centering \input{\thedatafolder/id_2.txt} &  {\scriptsize \input{\thedatafolder/abs1_2.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.18\textwidth]{\thedatafolder/img_1.pdf} & \centering \input{\thedatafolder/cycle_1.txt} \\ (2013) & \centering \input{\thedatafolder/id_1.txt} &  {\scriptsize \input{\thedatafolder/abs1_1.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.18\textwidth]{\thedatafolder/img_3.pdf} & \centering \input{\thedatafolder/cycle_3.txt} \\ (2016) & \centering \input{\thedatafolder/id_3.txt} &  {\scriptsize \input{\thedatafolder/abs1_3.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.18\textwidth]{\thedatafolder/img_0.pdf} & \centering \input{\thedatafolder/cycle_0.txt} \\ (2019) & \centering \input{\thedatafolder/id_0.txt} &  {\scriptsize \input{\thedatafolder/abs1_0.txt}} \tabularnewline
      \bottomrule
  \end{tabular}
  \caption{Examples of \hubble images (left-most column) and corresponding clipped proposal abstracts (right-most column). The observation cycle and corresponding year, as well as proposal ID are shown in the second and third columns, respectively. The proposal IDs link to the Mikulski Archive for Space Telescopes (MAST) page corresponding to the proposal.}
  \label{tab:dataset}
\end{table}

@@

The guided generation process ensures that, in this case, the output of the LLM strictly conforms to the format of the following example:
% \SM{Have it be centered}
% \begin{lstlisting}[language=Python]
% {
%   'objects_and_phenomena': ['star forming galaxy', 'lensed galaxy', ...], 
%   'science_use_cases': ['measure lensing magnification', 'probe spectral energy distributions', ...]
% }
% \end{lstlisting}
\begin{center}
  \begin{jsoncode}
    \centering
          \color{black}\{
            \color{codegreen}'objects_and_phenomena'\color{black}: [\color{codegreen}'star forming galaxy', 'lensed galaxy'\color{black}], 
            \color{codegreen}'science_use_cases'\color{black}: [\color{codegreen}'measure lensing magnification'\color{black}]
          \color{black}\}
  \end{jsoncode}
  \end{center}
which is then used to construct the summarized caption by combining the two key elements.
%
Examples of raw abstract snippets and corresponding LLM-generated summaries are shown in Tab.~\ref{tab:datasetsumm}.
%
% We emphasize that the intended goal of summarization-via-guided-generation is to increase the signal between text and images by standardizing the captions used for fine-tuning the CLIP model, and compare the quantitative performance of the model vs using the raw abstracts in Sec.~\ref{sec:results}.
%
We train separate models using the raw abstracts and the LLM-generated summaries, and compare their performance on downstream tasks in Sec.~\ref{sec:results}.
%
We note that, even after summarization, the association signal is expected to be noisy, since parts of the summarized caption may not be directly descriptive of the observed images. The goal of the fine-tuning process is to leverage the signal contained in this noisy association.

@@

CLIP \citep[Contrastive Language-Image Pre-training;][]{radford2021learning} is a multi-modal neural network model pre-trained on a large corpus of image-text pairs via weak supervision using a contrastive loss.
%
Given a minibatch $\mathcal{B}$ of $|\mathcal{B}|$ image-text pairs $\{(I_i, T_i)\}$, the goal is to align the learned representations of corresponding (positive) pairs $(I_i, T_i)$ while repelling the representations of unaligned (negative) pairs $(I_i, T_{j\neq i})$.
%
Image and text encoders $f: I \rightarrow \mathbb R^{n_\text{emb}}$ and $g: T \rightarrow \mathbb R^{n_\text{emb}}$ are used to map images and text to a common embedding space of dimension $n_\text{emb}$.
%
We use the standard softmax-based bidirectional variant of the InfoNCE~\citep{oord2018representation} contrastive loss function introduced for training CLIP-style architectures \citep{radford2021learning}
%
\begin{equation}
  \label{eq:softmax_loss}
  \mathcal{L}(\mathcal{B})=-\frac{1}{2|\mathcal{B}|} \sum_{i=1}^{|\mathcal{B}|}\left(\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_i \cdot y_j / \tau}}+\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_j \cdot y_i / \tau}}\right)
\end{equation}
%
where ${x}_i={f\left(I_i\right)}/{\left\|f\left(I_i\right)\right\|}$ and ${y}_i={g\left(T_i\right)}/{\left\|g\left(T_i\right)\right\|}$ are the normalized representations of the $i$-th image and text, respectively, and $\tau$ is a learnable temperature hyperparameter.
%
Note that this loss treats the image and text representations symmetrically, ensuring that the two modalities are considered on the same footing.

We use the CLIP-ViT-B/16\footnote{\url{https://huggingface.co/openai/clip-vit-base-patch16}} \citep{radford2021learning} variant as the base pre-trained CLIP model.
%
This model uses a 12-layer, 12-head, 768-embedding dimension vision transformer as the image encoder and a 12-layer, 8-head, 512-embedding dimension text sequence transformer as the text backbone.
%
The text encoder has a maximum length of 77 tokens and the image encoder a native resolution of $224\times224$ pixels.
%
Linear projection layers map the outputs of the image and text encoders to a common embedding space of dimension $n_\text{emb}=512$.
%
In total, the model has 149,620,737 trainable parameters.
%
This model was originally pre-trained on 400 million image-text pairs from internet data \citep{radford2021learning}.
%

@@

The base CLIP model is fine-tuned using the dataset described in Sec.~\ref{sec:dataset}, using either the LLM-summarized abstracts or raw proposal abstracts paired with observations.
%
When using raw proposal abstracts, random chunks of the text delimited by periods are selected on the fly to fit within the maximum token length of the text encoder.
%
Images are augmented via random four-fold rotations (increments of $90^\circ$) and randomly cropped to the native resolution of the image encoder, maintaining $\sim 20\%$ of the area of the original image, at each training step.
%
Given the relatively modest size of the fine-tuning dataset, a batch size $|\mathcal B| = 32$ is used throughout; larger batch sizes were observed to be susceptible to overfitting.
%
The temperature hyperparameter $\tau$ was initialized to its pre-trained value.
%
We emphasize that the positive and negative image-text association is noisy and imperfect, since multiple images can be associated with the same abstract, and the goal of the fine-tuning process is to leverage the signal contained in this noisy association.

@@

Figure~\ref{fig:sim_valtrain} (right) shows the retrieval accuracy, as defined in Eq.~\eqrefb{eq:retrieval_accuracy}, as a function of the retrieval fraction $k$.
%
In this case, we evaluate all four models (fine tuned on raw abstracts (blue), fine-tuned on LLM-summarized abstracts (red), trained on LLM-summarized abstracts from scratch (yellow), and the base model (purple)) on the same captions dataset -- the summarized abstracts -- for a direct comparison.
%
Remarkably, the model trained on raw abstracts shows very similar performance when evaluated on the summarized abstracts compared to that trained on the summarized abstracts themselves, indicating that \emph{(1)} the image-text association signal is preserved in the summarization process, and \emph{(2)} the model is able to effectively leverage meaningful concepts in the noisy raw abstracts through weak supervision.

@@

Having aligned the image and text representations, we can embed a natural language query using the model and show the closest images by embedding from the validation set when ranked by cosine similarity.
%
We show these in Tabs.~\ref{tab:tti_base} and \ref{tab:tti} for the base and fine-tuned models respectively using four simple curated queries: \texttt{dwarf galaxy} (small galaxies that typically orbit larger galaxies like the Milky Way), \texttt{Jupiter},  \texttt{SN1987A} (a specific, prominent supernova), and \texttt{strong lensing} (the phenomenon of bending of light due to the gravitational influence of a foreground distribution of matter). The proposal ID corresponding to the retrieved images is shown below each image, and contains a hyperlink to the MAST page corresponding to the proposal for further details.

@@

The model fine-tuned with summarized abstracts, meanwhile, shows strikingly different behavior (Tab.~\ref{tab:tti}).
%
% For example, it is able to return images with processing and assembly artifacts particular to HST (e.g., the lines through the middle in some images), which typically receive low similarity scores when using the base model.
%
The \texttt{dwarf galaxy}-queried images correspond to proposals aiming to measure the kinematics of the stellar cores of dwarf galaxies.
%
Images of Jupiter are returned for the \texttt{Jupiter} query, with the planet clearly visible in all four images.
%
Supernova SN1987 itself can be seen in the three closest images for the \texttt{SN1987A} query with the fourth image being a supernova remnant.
%
Cluster-scale as well as galaxy-scale gravitational lenses are returned by the \texttt{strong lensing} query, with lensing patterns visible in the images.

@@

The second row (an image of supernova 1987A) highlights an interesting pattern -- the base model erroneously attributes the object at the center of the image to a gravitational lens, while the fine-tuned model correctly identifies it as a supernova remnant. The kind of reasonable misattribution is common when querying the base model, and largely absent in the fine-tuned model.

Note that we chose to illustrate qualitative performance on text and image retrieval using the model fine-tuned on summarized abstracts, rather than raw abstracts. We show analogous results for the model fine-tuned on raw abstracts in App.~\ref{app:eval_raw}. Although the two models show very similar quantitative performance on retrieval metrics (as shown in Fig.~\ref{fig:sim_valtrain}), they exhibit characteristically different behaviors in terms of objects (images/text) retrieved. We emphasize that for scientific usefulness, the goal is not necessarily to correctly retrieve the most ``relevant'' objects, but rather to identify a diverse set of interesting candidates for manual follow-up and further analysis; both models are seen to perform sensibly, even if differently, in this regard.

@@

\begin{table}[h!]
  \centering
  \renewcommand{\arraystretch}{0.1}
  \begin{tabular}{m{3cm} m{3.6cm} m{3.6cm} m{4cm}}
      \toprule
      \centering \bfseries \hubble image & \centering \textbf{Top-4 text} \\ \small\textbf{\textcolor{deeppurple}{(base)}} & \centering  \textbf{Top-4 text} \\ \small\textbf{\textcolor{deepred}{(summary fine-tuned)}} & \centering \textbf{Summarized abstract} \\ (objects; `ground truth') \tabularnewline
      \midrule
      \centering \includegraphics[width=0.15\textwidth]{\thedatafolder/img_itt_0.pdf} & \centering \scriptsize \verbatiminput{\thedatafolder/sci_itt_base_0.txt} & \centering  \scriptsize \verbatiminput{\thedatafolder/sci_itt_0.txt} &  {\scriptsize \input{\thedatafolder/abs_itt_0.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.15\textwidth]{\thedatafolder/img_itt_1.pdf} & \centering \scriptsize \verbatiminput{\thedatafolder/sci_itt_base_1.txt} & \centering  \scriptsize \verbatiminput{\thedatafolder/sci_itt_1.txt} &  {\scriptsize \input{\thedatafolder/abs_itt_1.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.15\textwidth]{\thedatafolder/img_itt_2.pdf} & \centering \scriptsize \verbatiminput{\thedatafolder/sci_itt_base_2.txt} & \centering  \scriptsize \verbatiminput{\thedatafolder/sci_itt_2.txt} &  {\scriptsize \input{\thedatafolder/abs_itt_2.txt}} \tabularnewline
      \midrule
      \centering \includegraphics[width=0.15\textwidth]{\thedatafolder/img_itt_3.pdf} & \centering \scriptsize \verbatiminput{\thedatafolder/sci_itt_base_3.txt} & \centering  \scriptsize \verbatiminput{\thedatafolder/sci_itt_3.txt} &  {\scriptsize \input{\thedatafolder/abs_itt_3.txt}} \tabularnewline
      \bottomrule
  \end{tabular}
  \caption{Text snippets from a curated list most closely matching a given image query (left-most column) by cosine similarity of respective embeddings, shown for the \textcolor{deeppurple}{base} (CLIP-ViT-B/16) and \textcolor{deepred}{summary fine-tuned} models. The `ground truth' LLM-summarized abstract (only objects/phenomena) is shown in the right-most column.}
  \label{tab:itt}
\end{table}

@@

In this paper we present \textsc{PAPERCLIP}, a method for training domain-specific multi-modal models for astrophysics that associates observations imaged by telescopes with natural language in a common, semantically-meaningful embedding space.
%
We showcase an application to \hubble Space Telescope (HST) observations, where the model is fine-tuned from a pre-trained CLIP model using abstracts of successful \hubble proposals, optionally summarized, leveraging a noisy association signal between text and images.
%
We show that \textsc{PAPERCLIP} significantly outperforms the base CLIP model in quantitative metrics, such as retrieval accuracy, as well as quality of text-to-image and image-to-text retrieval.
%
We also introduce a novel LLM summarization process which leverages guided generation to distill the content of proposal abstracts while preserving salient information. 
%
Overall, the procedure demonstrates the efficacy of fine-tuning generalist pre-trained models on small amounts of domain-specific data, in particular astronomical datasets, and leveraging text as an interface.
